CTL mapping - The added value from segregating correlation

### 2012-02-09 inst/methodPaper/document.tex
Document

@@

\begin{document}
\firstpage{1}
\title[CTLmapping]{CTL mapping - The added value from segregating correlation}
\author[Arends \emph{et al.}]{
  Danny Arends\,$^{1,\footnote{to whom correspondence should be addressed}}$,
  and Ritsert C. Jansen\,$^{1}$
}
\address{
  $^{1}$Groningen Bio-informatics Centre, University of Groningen, Groningen, The Netherlands.
}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXX}
\maketitle
\begin{abstract}
  \section{Motivation:}
  Here we present Correlated Traits Locus (CTL) mapping, a novel approach 
  to detect genetic regulation of traits in natural and inbred populations. 
  CTL mapping allows association of trait correlation differences to genetic 
  markers, in almost the same way as QTL mapping associates genetic  markers with differences in expression.

  We use CTL mapping to reconstruct underlying genetic architecture, and provide insight 
  into regulation of (gene) expression. CTL information in combination with classical QTL 
  information leads to a more detailed insight into the genetic architecture underlying 
  complex traits. CTL mapping provides a missing link in regulatory network reconstruction.

  \section{Results:}
  The software is available under the terms of GNU General Public License, version 3
  (\href{http://www.r-project.org/Licenses/GPL-3}{www.r-project.org/Licenses/GPL-3}).
  \section{Availability:}
  Software, documentation and source code are available at request.
  \section{Contact:} 
  \href{Danny.Arends@Gmail.com}{Danny.Arends@Gmail.com}
\end{abstract}

@@

  \begin{figure*}[ht]
    \begin{center}
    \vspace{-0.2cm}
    \hspace{-1.3cm}
    \subfigure[CTL profiles]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.eps}
      \label{subfiga}
    }
    \hspace{0.3cm}
    \subfigure[Traits x Traits interaction matrix]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.eps}
      \label{subfigb}
    }
    \hspace{0.3cm}
    \subfigure[Reconstructed network]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_net.eps}
      \label{subfigc}
    }
  \end{center}
  \begin{minipage}{7in}
  \label{newplots}
  \vspace{-0.7cm}
  \caption[Plots]{
  {\emph {\bf a}}: Example CTL + QTL profiles. Seen are the QTL profile on top of the CTL profile. 
  The first peak is a CTL, we don't observe a QTL for this location. The second peak is a QTL peek, with no underlying CTL. 
  {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
  showing for each trait the significance of the difference in correlation to all other traits. This 
  matrix can be transformed into .sif file to generate a network image. 
  {\emph {\bf c}}:  Reconstructed network, this shows the same information as the TxT matrix only in a more accessible way, we visualized our 
  network using cytoscape, the width of the line is the maximum interaction evidence.
 }
\end{minipage}
\vspace{-0.5cm}
\end{figure*}

@@

\emph{ {\bf Materials}}
  Metabolite data from an \emph{A. Thaliana} Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
  dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
  There are 162 recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
  details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}
\vspace{-0.6cm}
\section{Results}
  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was 
  determined by using 15,000 permutations.
  When combining CTL and QTL information, we observe:\\
  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - P1). Loss of variation in the genes 
  underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
  detect the differences in correlation.\\
  2) QTL with no CTL (Fig \ref{subfiga} - P2), we interpret this as the source of the variation. Analysis in this way of 
  all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
  source of natural variation and follow it down the network.
  3) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
  are contributing variation to the observed QTL  (Fig \ref{subfiga} - P3).\\

@@

\section{Conclusion \& Discussion}
  Microarray experiments have identified genes differentially expressed between various 
  conditions, or groups of genes coexpressed across samples. Genetical genomics screens 
  are carried out to map differential expression to genetic variation. Screening 
  the genome for genotype combinations for which there is differential expression is 
  a successful strategy when mapping Mendelian traits.

>>>>

\usepackage{color,soul}  % for markup
\setul{1ex}{0.8ex}
\definecolor{orange}{rgb}{1,0.5,0}
\newcommand{\marked}[1]{\hl{#1}}
\newcommand{\comment}[1]{\hl{\bf FIXME! #1}}

@@

\begin{document}
\firstpage{1}
\title[CTLmapping]{CTL mapping - The added value from segregating correlation}

\author[Arends \emph{et al.}]{
  Danny Arends\,$^{1,\footnote{to whom correspondence should be addressed}}$,
  Pjotr Prins\,$^{1,2}$
  and Ritsert C. Jansen\,$^{1}$
}
\address{
  $^{1}$Groningen \hl{Bio-informatics} Centre, University of Groningen, Groningen, The Netherlands.
  $^{2}$Laboratory of Nematology, Wageningen University, The Netherlands.
}
\history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}
\editor{Associate Editor: XXXXX}
\maketitle

\comment{Title is too cute, and looks too much like an earlier paper;
also should be more descriptive}

\comment{3x Groningen in address. And does that dash belong there?}

\begin{abstract}
  \section{Summary:}

  % Summary moet 1 of 2 zinnen zijn! Zoiets beter?

  Mapping of correlated traits loci (CTL) to the genome potentially
  allows the detection of genetic regulation of traits by tracking
  correlation differences at genetic marker locations. CTL can be
  combined with quantitative trait loci (QTL) in regulatory network
  reconstruction.

  %  , in almost the same way as QTL mapping associates genetic
  %  markers with differences in expression.  <- als je dit zegt is
  %  het niet zo nieuw meer ;)

  % We use CTL mapping to reconstruct underlying genetic architecture, and provide insight 
  % into regulation of (gene) expression. CTL information in combination with classical QTL 
  % information leads to a more detailed insight into the genetic architecture underlying 
  % complex traits. CTL mapping

  \section{Availability:}

  The software is available under the terms of GNU General Public License, version 3
  (\href{http://www.r-project.org/Licenses/GPL-3}{www.r-project.org/Licenses/GPL-3}).
  \section{Availability:}
  Software, documentation and source code are available at request.
  \section{Contact:} 
  \href{Danny.Arends\hl{@Gmail.com}}{Danny.Arends@\hl{Gmail.com}}
\end{abstract}

\comment{Use university E-mail}

\comment{License: BSD license is preferred these days}

\comment{Availability: Bioinformatics Journal wants open source software only,
availability should reflect that - i.e. a github.org repo. They won't
like your option ;)}

@@

  \begin{figure*}[ht]
    \begin{center}
    \vspace{-0.2cm}
    \hspace{-1.3cm}
    \subfigure[CTL profiles]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_qtl.eps}
      \label{subfiga}
    }
    \hspace{0.3cm}
    \subfigure[Traits x Traits interaction matrix]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_pxp.eps}
      \label{subfigb}
    }
    \hspace{0.3cm}
    \subfigure[Reconstructed network]{
      \includegraphics[keepaspectratio,scale=.3]{img/ctl_net.eps}
      \label{subfigc}
    }
  \end{center}
  \begin{minipage}{7in}
  \label{newplots}
  \vspace{-0.7cm}
  \caption[Plots]{
  {\emph {\bf a}}: Example CTL + QTL profiles. Seen are the QTL profile on top of the CTL profile. 
  The first peak is a CTL, we don't observe a QTL for this location.
  The second peak is a QTL peek, with no underlying CTL. \hl{You
  should explain P1,P2 and P3 in the caption too! A figure should be
  read on its own} 
  {\emph {\bf b}}: Traits x Traits interaction matrix (TxT matrix). Information obtained using CTL mapping, 
  showing for each trait the significance of the difference in correlation to all other traits. This 
  matrix can be transformed into .sif file to generate a network image. 
  {\emph {\bf c}}:  Reconstructed network, this shows the same information as the TxT matrix only in a more accessible way, we visualized our 
  network using cytoscape, the width of the line is the maximum interaction evidence.
 }
\end{minipage}
\vspace{-0.5cm}
\end{figure*}

@@

\emph{ {\bf Materials}}
  Metabolite data from an \emph{A. Thaliana} Landsberg erecta (Ler) and Cape Verde Islands (Cvi) 
  dataset from R/qtl, 24 characterized metabolites which are the products of chain elongation. 
  There are 162 recombinant inbred lines genotyped at 117 markers on 5 chromosomes. Additional 
  details about the experimental settings are available in the original paper \cite{Keurentjes:2006,AlonsoBlanco:1998}
\vspace{-0.6cm}
\section{FEATURES}
  CTL mapping was performed and a per trait CTL thresholds of $P < 0.001$ was 
  determined by using 15,000 permutations.
  When combining CTL and QTL information, we observe:\\
  1) Additional CTL found, where no QTL was present (Fig \ref{subfiga} - P1). Loss of variation in the genes 
  underneath the CTL are not causing variation differences in the observed trait, CTL is able to 
  detect the differences in correlation.\\
  2) QTL with no CTL (Fig \ref{subfiga} - P2), we interpret this as the source of the variation. Analysis in this way of 
  all 24 Single trait CTL profiles combined with their QTL profiles allows us to reconstruct the 
  source of natural variation and follow it down the network.
  3) CTL co-localized with QTL, in this situation CTL gives a detailed view of which traits 
  are contributing variation to the observed QTL  (Fig \ref{subfiga} - P3).\\

@@

\section{Conclusion}
  Microarray experiments have identified genes differentially expressed between various 
  conditions, or groups of genes coexpressed across samples. Genetical genomics screens 
  are carried out to map differential expression to genetic variation. Screening 
  the genome for genotype combinations for which there is differential expression is 
  a successful strategy when mapping Mendelian traits.
